00-Data-Processing-for-Food-Banks-Accessibility

### 2021-04-21 00-Data-Processing/00-Data-Processing-for-Food-Banks-Accessibility.Rmd
Add population output
Clean up pre-processing code and add output by population groups from tts

### Hamilton CMA boundaries

Statistics Canada. Load:
```{r}
load("input-data-files/hamilton_cma.RData")
#load("input-data-files/hamilton_da_2016.RData")
```

### Hamilton Ward boundaries

Open Data Hamilton. Read:
```{r}
wards <- st_read("input-data-files/Ward_Boundaries.shp")
```

Add label for type of ward (the urban/suburban/rural classification is by the planning teams; see https://www.hamilton.ca/develop-property/planning-applications/development-applications-mapping)
```{r}
wards$Type <- c("Urban",
                "Urban",
                "Urban",
                "Urban",
                "Urban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban",
                "Suburban")
```

Merge boundaries by type:
```{r}
wards <- wards %>% 
  group_by(Type) %>%
  summarize()
```

```{r}
ggplot(wards) + 
  geom_sf(aes(fill = Type))
```

Read rural boundary:
```{r}
rural <- st_read("input-data-files/Rural_Boundary.shp")
rural <- rural %>%
  transmute(Type = "Rural")
```

Plot rural boundary:
```{r}
ggplot(rural) +
  geom_sf()
```

Obtain difference between urban and suburban regions and rural boundary:
```{r}
st_difference(wards, rural) %>%
  ggplot() +
  geom_sf(aes(fill = Type))
```

Notice the detritus after the difference. Drop crumbs to obtain a cleaner version of the boundaries:
```{r}
sub_urban <- st_difference(wards, rural) %>%
  transmute(Type) %>%
  smoothr::drop_crumbs(threshold = 5000)

ggplot(sub_urban) +
  geom_sf(aes(fill = Type))
```

Bind the boundaries with all three urban types:
```{r}
urban_types <- rbind(sub_urban,
                     rural)

ggplot(urban_types) +
  geom_sf(aes(fill = Type))
```

@@

Convert to simple features:
```{r message = FALSE}
foodbank_locations <- foodbank_locations %>%
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = 4326) %>%
  #st_transform(crs = 26917) %>%

  # filter out points beyond the Hamilton CMA (there's one point in Brantford)
  filter(st_intersects(x = geometry, y = hamilton_cma, sparse = FALSE))
```

Map location of foodbanks:
```{r}
ggplot() +
  geom_sf(data = hamilton_cma) +
  geom_sf(data = foodbank_locations,
          aes(shape = COVIDClose,
              color = COVIDClose))
```

@@

```{r}
load("input-data-files/dwelling_network_points_2016.RData")
```

Use `cancensus` to check and download census data. Check the regions:
```{r}
list_census_regions('CA16') %>% 
  dplyr::filter(level == "CMA", name %in% c("Hamilton"))
```

Search variables:
```{r}
find_census_vectors("age", 
                    dataset = "CA16", 
                    type = "total", 
                    query_type = "keyword", 
                    interactive = FALSE)
```

We need v_CA16_2406 to v_CA16_2425: these are the number of households at certain levels of total income. Retrieve data:
```{r}
data_da_2016 <- get_census(dataset='CA16', 
                          regions=list(CMA=c("35537")),
                          vectors=c("v_CA16_2406",
                                    "v_CA16_2407",
                                    "v_CA16_2408",
                                    "v_CA16_2409",
                                    "v_CA16_2410",
                                    "v_CA16_2411",
                                    "v_CA16_2412",
                                    "v_CA16_2413",
                                    "v_CA16_2414",
                                    "v_CA16_2415",
                                    "v_CA16_2416",
                                    "v_CA16_2417",
                                    "v_CA16_2418",
                                    "v_CA16_2419",
                                    "v_CA16_2420",
                                    "v_CA16_2421",
                                    "v_CA16_2422",
                                    "v_CA16_2423",
                                    "v_CA16_2424",
                                    "v_CA16_2425",
                                    "v_CA16_2426",
                                    "v_CA16_2397", # median hhinc
                                    # age variables
                                    "v_CA16_10", #<1
                                    "v_CA16_13", #1
                                    "v_CA16_16", #2
                                    "v_CA16_7", #0-4 yo
                                    "v_CA16_25", #5-9 yo
                                    "v_CA16_43", #10-14 yo
                                    "v_CA16_64", # 15-19 yo
                                    "v_CA16_244"), # 65+
                          level='DA', 
                          use_cache = FALSE,
                          geo_format = "sf")
```

Clip `data_da_2016`:
```{r}
data_da_2016 <- data_da_2016 %>%
  st_intersection(hamilton_cma)
```

Calculate number of households with incomes < \$40k per DA and proportion of households with incomes < \$40k per DA:
```{r}
data_da_2016 <- data_da_2016 %>%
  filter(`Region Name` == "Hamilton") %>%
  mutate(income_less40k = (`v_CA16_2406: Under $5,000` + 
                             `v_CA16_2407: $5,000 to $9,999` + 
                             `v_CA16_2408: $10,000 to $14,999` +
                             `v_CA16_2409: $15,000 to $19,999` +
                             `v_CA16_2410: $20,000 to $24,999` +  
                             `v_CA16_2411: $25,000 to $29,999` +
                             `v_CA16_2412: $30,000 to $34,999` +
                             `v_CA16_2413: $35,000 to $39,999`),
         p_inc_less40k = income_less40k/Households)
```

Plot population:
```{r}
tmap_mode(mode = "view")

data_da_2016 %>%
  tm_shape() +
  tm_polygons("income_less40k",
              alpha = 0.8)
```

Calculate low-income age totals
```{r}
data_da_2016 <- data_da_2016 %>%
  mutate(tot_age_0to2 = `v_CA16_10: Under 1 year` + `v_CA16_13: 1` + `v_CA16_16: 2`,
         linc_age_0to2 = tot_age_0to2 * p_inc_less40k,
         tot_age_0to19 = `v_CA16_7: 0 to 4 years` + `v_CA16_25: 5 to 9 years` + `v_CA16_43: 10 to 14 years` + `v_CA16_64: 15 to 19 years`,
         linc_age_0to19 = tot_age_0to19 * p_inc_less40k,
         tot_age_65plus = `v_CA16_244: 65 years and over`,
         linc_age_65plus = tot_age_65plus * p_inc_less40k,
         linc_population = Population * p_inc_less40k)
```

Plot result
```{r}
tmap_mode(mode = "view")

data_da_2016 %>%
  tm_shape() +
  tm_polygons("linc_population",
              alpha = 0.8)
```

@@

Read the shape file with the traffic analysis zone boundaries:
```{r}
tts_taz <- st_read("input-data-files/tts2016_2006zn_2018_region.shp")
```

@@

Save data files:
```{r}
save(data_da_2016, file = "output-data-files/data_da_2016.RData", compress = "xz")
save(dwelling_network_points_2016, file = "output-data-files/dwelling_network_points_2016.RData", compress = "xz")
save(foodbank_locations, file = "output-data-files/foodbank_locations.RData", compress = "xz")
save(hamilton_cma, file = "output-data-files/hamilton_cma.RData", compress = "xz")
save(modes_less40k, file = "output-data-files/modes_less40k.RData", compress = "xz")
save(population_age_income, file = "output-data-files/population_age_income.RData", compress = "xz")
save(urban_types, file = "output-data-files/urban_types.RData", compress = "xz")
```

>>>>

### Census Data

First, load the water boundary for Hamilton Harbour:

```{r}
load("input-data-files/hamilton_harbour.RData")
```

Use `cancensus` to check and download census data. Check the regions:
```{r message = FALSE}
list_census_regions('CA16') %>% 
  dplyr::filter(level == "CSD", name %in% c("Hamilton"))
```

Search variables for DA-level data:
```{r}
find_census_vectors("age", 
                    dataset = "CA16", 
                    type = "total", 
                    query_type = "keyword", 
                    interactive = FALSE)
```

We need v_CA16_2406 to v_CA16_2425: these are the number of households at certain levels of total income. Retrieve data:
```{r}
data_da_2016 <- get_census(dataset='CA16', 
                          regions=list(CSD=c("3525005")),
                          vectors=c("v_CA16_2406",
                                    "v_CA16_2407",
                                    "v_CA16_2408",
                                    "v_CA16_2409",
                                    "v_CA16_2410",
                                    "v_CA16_2411",
                                    "v_CA16_2412",
                                    "v_CA16_2413",
                                    "v_CA16_2414",
                                    "v_CA16_2415",
                                    "v_CA16_2416",
                                    "v_CA16_2417",
                                    "v_CA16_2418",
                                    "v_CA16_2419",
                                    "v_CA16_2420",
                                    "v_CA16_2421",
                                    "v_CA16_2422",
                                    "v_CA16_2423",
                                    "v_CA16_2424",
                                    "v_CA16_2425",
                                    "v_CA16_2426",
                                    "v_CA16_2397"), # median hhinc
                          level='DA', 
                          use_cache = FALSE,
                          geo_format = "sf") %>%
  # project to NAD 1983 Zone 17N
  st_transform(crs = 26917)
```

Clip `data_da_2016` to remove the Hamilton Harbour from the CSD polygons:
```{r warning = FALSE}
data_da_2016 <- st_difference(data_da_2016, st_union(st_combine(hamilton_harbour %>% filter(NAME == "Hamilton Harbour")))) %>% drop_crumbs(threshold = 1)

tm_shape(data_da_2016) + 
  tm_fill(col = "Population")
```

Create a Hamilton CSD (municipal) boundary polygon:
```{r}
hamilton_boundary <- data_da_2016 %>%
  group_by(CSD_UID) %>%
  summarize(Population = sum(Population))

tm_shape(hamilton_boundary) + 
  tm_fill(col = "Population")
```

Calculate number of households with incomes < \$40k per DA and proportion of households with incomes < \$40k per DA:
```{r}
data_da_2016 <- data_da_2016 %>%
  #filter(`Region Name` == "Hamilton") %>%
  mutate(income_less40k = (`v_CA16_2406: Under $5,000` + 
                             `v_CA16_2407: $5,000 to $9,999` + 
                             `v_CA16_2408: $10,000 to $14,999` +
                             `v_CA16_2409: $15,000 to $19,999` +
                             `v_CA16_2410: $20,000 to $24,999` +  
                             `v_CA16_2411: $25,000 to $29,999` +
                             `v_CA16_2412: $30,000 to $34,999` +
                             `v_CA16_2413: $35,000 to $39,999`),
         p_inc_less40k = income_less40k/Households)
```

Plot population:
```{r}
tm_shape(data_da_2016) + 
  tm_fill(col = "income_less40k", alpha = 0.8)
```

```{r income-quintiles, include=FALSE}
# Data preparation
## Calculate quintiles of median household income in Hamilton:
mhi_q <- data_da_2016 %>%
  dplyr::filter(`Region Name` == "Hamilton") %>%
  pull(`v_CA16_2397: Median total income of households in 2015 ($)`) %>%
  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), 
           na.rm = TRUE)
```

@@

Convert to simple features:
```{r message = FALSE}
foodbank_locations <- foodbank_locations %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 26917) %>% # project to NAD 1983 Zone 17N

  # filter out points beyond the Hamilton CMA (there's one point in Brantford)
  filter(st_intersects(x = geometry, y = hamilton_boundary, sparse = FALSE))
```

Map location of foodbanks:
```{r}
ggplot() +
  geom_sf(data = hamilton_boundary) +
  geom_sf(data = foodbank_locations,
          aes(shape = COVIDClose,
              color = COVIDClose))
```

@@

```{r}
load("input-data-files/dwelling_network_points_2016.RData")

dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
  st_transform(crs = 26917) # project to NAD 1983 Zone 17N
```

@@

Read the shape file with the traffic analysis zone boundaries:
```{r}
tts_taz <- st_read("input-data-files/tts2016_2006zn_2018_region.shp") %>% 
  st_transform(crs = 26917) %>% # project to NAD 1983 Zone 17N
  st_make_valid() # fix a few topolgy errors in the boundaries
```

@@

Save data files:
```{r}
save(data_da_2016, file = "output-data-files/data_da_2016.RData", compress = "xz")
save(hamilton_boundary, file = "output-data-files/hamilton_boundary.RData", compress = "xz")
save(dwelling_network_points_2016, file = "output-data-files/dwelling_network_points_2016.RData", compress = "xz")
save(foodbank_locations, file = "output-data-files/foodbank_locations.RData", compress = "xz")
save(modes_less40k, file = "output-data-files/modes_less40k.RData", compress = "xz")
save(population_age_income, file = "output-data-files/population_age_income.RData", compress = "xz")
#save(urban_types, file = "output-data-files/urban_types.RData", compress = "xz")
```
